---
title: "Data Visualization in R"
subtitle: "EPI 553 | Principles of Statistical Inference II"
author: "Muntasir Masum, PhD"
date: "April 21 & 24, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    transition: none
    transition-speed: fast
    background-transition: none
    slide-number: c/t
    chalkboard: true
    preview-links: auto
    code-line-numbers: false
    code-fold: false
    code-overflow: wrap
    highlight-style: a11y-light
    height: 900
    width: 1600
    fig-align: center
    fig-format: svg
    footer: "EPI 553 | Data Visualization | University at Albany"
    logo: img/ualbany-logo.png
    title-slide-attributes:
      data-background-color: "#0e1c36"
      data-background-image: "https://images.unsplash.com/photo-1551288049-bebda4e38f71?w=1600&q=80"
      data-background-opacity: "0.15"
    navigation-mode: linear
    controls: true
    mouse-wheel: false
    pointer:
      pointerSize: 18
      color: "#e64173"
execute:
  echo: true
  warning: false
  message: false
  fig-width: 12
  fig-height: 6.5
  fig-dpi: 150
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(palmerpenguins)
library(knitr)
library(kableExtra)
library(patchwork)
library(scales)

# Consistent EPI 553 theme
theme_epi553 <- function(base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 4,
                                color = "#1a1a2e", margin = margin(b = 4)),
      plot.subtitle = element_text(color = "grey40", size = base_size,
                                   margin = margin(b = 12)),
      plot.caption = element_text(color = "grey60", size = base_size - 3,
                                  hjust = 0, margin = margin(t = 12)),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      axis.title = element_text(color = "grey30"),
      axis.text = element_text(color = "grey30"),
      strip.text = element_text(face = "bold", color = "grey20"),
      legend.position = "top",
      legend.title = element_text(face = "bold", size = base_size - 1),
      plot.title.position = "plot",
      plot.caption.position = "plot"
    )
}

# Set as default
theme_set(theme_epi553())

# Palette
epi_colors <- c("Adelie" = "#FF6B35", "Chinstrap" = "#A23B72", "Gentoo" = "#2E86AB")

set.seed(1220)
```


## Roadmap {.smaller}

:::: {.columns}

::: {.column width="50%"}

### Foundations

1. Why visualization is a methods topic
2. The grammar of graphics
3. Healy's principles: honesty, clarity, comparison
4. Holtz's chart-type decision tree
5. Scherer's editorial approach
6. A worked transformation: same data, four iterations

:::

::: {.column width="50%"}

### Modern Techniques

7. Color, type, and accessibility
8. gghighlight, animation, interactivity
9. Distributions, uncertainty, heatmaps
10. Patchwork, annotation, spatial viz
11. **Visualizing regression models**
12. The #30DayChartChallenge
13. Building your own reusable theme

:::

::::


# Part 1 {background-color="#46166B" .section-slide}

## Why Visualization Is a Methods Topic

Statistics gives us numbers. Visualization makes them mean something.


## Three Reasons

::: {.incremental}

1. **Visualization is part of the analysis, not decoration.**
   A scatterplot reveals a non-linear relationship that a correlation coefficient hides. A residual plot exposes a violated assumption that an R-squared celebrates.

2. **Visualization is how findings reach decision-makers.**
   A clinician, journalist, or policymaker will not read your beta coefficients. They *will* look at your figure. The figure is the one part of the paper that everyone reads.

3. **Visualization is a discipline with its own theory.**
   Bad charts are not just ugly, they are *wrong*. They make comparisons hard, hide variation, and encode noise as signal.

:::

::: {.fragment}
> *"Above all else, show the data."* -- Edward Tufte
:::


## The Same Data, Completely Different Stories

```{r}
#| label: fig-anscombe
#| fig-cap: "Anscombe's Quartet: four datasets with identical summary statistics but radically different patterns."
#| echo: false
#| fig-height: 5

anscombe_long <- anscombe |>
  pivot_longer(everything(),
               names_to = c(".value", "set"),
               names_pattern = "(.)(.)")

ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 3, color = "#2E86AB", alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "#e64173", linewidth = 1) +
  facet_wrap(~set, ncol = 4, scales = "free") +
  labs(title = "Anscombe's Quartet",
       subtitle = "Same mean, same variance, same correlation, same regression line. Very different data.") +
  theme_epi553(base_size = 13)
```

::: {.muted}
Every analysis you do should **begin and end** with looking at the data.
:::


# Part 2 {background-color="#46166B" .section-slide}

## The Grammar of Graphics

Leland Wilkinson (1999) described a unified framework. Hadley Wickham translated it into ggplot2 (2005).


## The Seven Layers {.smaller}

:::: {.columns}

::: {.column width="50%"}

### Core structure

1. **Data**
   - What data am I plotting?
   - `ggplot(data = ...)`
2. **Aesthetics**
   - Which variables map to x, y, color, size?
   - `aes(x, y, color, size)`
3. **Geometries**
   - What shape do I draw?
   - `geom_point()`, `geom_line()`, `geom_col()`
4. **Facets**
   - Do I split into small multiples?
   - `facet_wrap()` / `facet_grid()`

:::

::: {.column width="50%"}

### Refinement layers

5. **Statistics**
   - Do I summarize or transform?
   - `stat_*()` / `geom_smooth()`
6. **Coordinates and scales**
   - How are axes and scales defined?
   - `coord_*()` / `scale_*()`
7. **Theme**
   - How does it look?
   - `theme_*()` / `theme()`

::: {.highlight-box}
Think of these as a checklist: data, mapping, marks, splits, summaries, scales, polish.
:::

:::

::::

::: {.fragment .highlight-box}
The genius is **composability**: to go from a scatterplot to a faceted scatterplot with a smoother, you *add* layers; you do not start over.
:::


## Building a Plot Layer by Layer

```{r}
#| label: fig-grammar-demo
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5.5

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~ island) +
  scale_color_manual(values = epi_colors) +
  labs(title = "Bill Dimensions of Palmer Penguins",
       subtitle = "By species and island, with linear trend lines",
       x = "Bill length (mm)", y = "Bill depth (mm)",
       color = "Species")
```


## The ggplot2 Template

Every ggplot follows the same skeleton:

```{r}
#| eval: false
#| code-line-numbers: "1|2|3|4|5|6|7"

ggplot(data = <DATA>,                       # 1. Data
       aes(x = <X>, y = <Y>, color = <Z>)) # 2. Aesthetics
  + geom_<TYPE>(...)                        # 3. Geometry
  + facet_wrap(~ <VAR>)                     # 4. Facets
  + scale_<AES>_<TYPE>(...)                 # 5. Scales
  + labs(title = "...", x = "...", y = "...") # 6. Labels
  + theme_minimal()                         # 7. Theme
```

::: {.fragment .finding-box}
You will use this template for every chart in this course and beyond.
:::


# Part 3 {background-color="#46166B" .section-slide}

## Kieran Healy's Principles

Is the chart **substantively** good? **Perceptually** good? **Aesthetically** good?


## Substantive Standards {.smaller}

:::: {.columns}

::: {.column width="40%"}

### The Rules

::: {.incremental}
- **Show the data.** Individual observations, not just summaries
- **Compare like with like.** Stratify intentionally
- **Quantify uncertainty.** Always show CIs or standard errors
:::

:::

::: {.column width="60%"}

```{r}
#| label: fig-healy-substantive
#| echo: false
#| fig-height: 6

penguins |>
  drop_na(sex) |>
  ggplot(aes(x = species, y = body_mass_g, color = sex)) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1.5) +
  stat_summary(fun.data = mean_cl_normal, geom = "pointrange",
               position = position_dodge(width = 0.5),
               size = 0.8, linewidth = 1) +
  scale_color_manual(values = c("female" = "#D55E00", "male" = "#0072B2")) +
  labs(title = "Body Mass by Species and Sex",
       subtitle = "Individual observations with mean and 95% CI",
       x = NULL, y = "Body mass (g)", color = "Sex")
```

:::

::::


## Perceptual Standards

Cleveland & McGill (1984) ranked visual encodings by accuracy:

```{r}
#| label: fig-perception
#| echo: false
#| fig-height: 5

perception <- tibble::tibble(
  encoding = c("Position (common scale)", "Position (non-aligned)",
               "Length", "Angle / slope", "Area",
               "Volume", "Color hue", "Color saturation"),
  accuracy = c(95, 82, 70, 55, 42, 30, 25, 18),
  rank = 1:8
)

ggplot(perception, aes(x = accuracy,
                       y = fct_reorder(encoding, accuracy))) +
  geom_col(fill = "#2E86AB", alpha = 0.85, width = 0.7) +
  geom_text(aes(label = paste0("#", rank)), hjust = -0.3, fontface = "bold",
            color = "#46166B", size = 5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = label_number(suffix = "%")) +
  labs(title = "Cleveland-McGill Hierarchy of Visual Encodings",
       subtitle = "Higher = more accurately perceived by readers",
       x = "Perceptual accuracy (approximate)", y = NULL)
```


## Practical Implication {.smaller}

::: {.highlight-box}
Prefer **dot plots** and **bar charts** to pie charts. Prefer **faceting** to stacking. Use color for categories, not for quantities (unless you use a perceptually uniform scale like viridis).
:::

:::: {.columns}

::: {.column width="50%"}

### Weak encoding (angle)
```{r}
#| echo: false
#| fig-height: 4.5

dat <- tibble(group = LETTERS[1:6],
              value = c(22, 18, 17, 16, 14, 13))

ggplot(dat, aes(x = "", y = value, fill = group)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pie chart: which is third largest?") +
  theme_void(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

:::

::: {.column width="50%"}

### Strong encoding (position)
```{r}
#| echo: false
#| fig-height: 4.5

ggplot(dat, aes(x = value, y = fct_reorder(group, value))) +
  geom_col(fill = "#2E86AB", width = 0.7) +
  geom_text(aes(label = value), hjust = -0.3, fontface = "bold", size = 5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Bar chart: instantly clear", x = NULL, y = NULL)
```

:::

::::


## Aesthetic Standards

Healy's third pillar is what beginners notice last but readers notice first:

::: {.incremental}
- **Typography**: one font family, bold title, grey subtitle, tiny caption
- **White space**: let the chart breathe
- **Alignment**: consistent margins and axes
- **Color harmony**: restrained, purposeful palettes
- **Direct labeling**: labels next to the data, not in a legend
:::

::: {.fragment .finding-box}
A clean theme, restrained gridlines, and direct labeling will make a competent chart feel **professional**.
:::


# Part 4 {background-color="#46166B" .section-slide}

## Choosing the Right Chart Type

Yan Holtz's *From Data to Viz* decision tree


## Chart Type Decision Guide {.smaller}

:::: {.columns}

::: {.column width="50%"}

### Start with the data structure

- **One numeric**
  Histogram, density, boxplot, violin
- **One categorical**
  Bar chart, lollipop, treemap
- **Two numeric**
  Scatterplot, hexbin, 2D density
- **Numeric x categorical**
  Boxplot, violin, ridge, jitter + summary

:::

::: {.column width="50%"}

### Then ask what to avoid

- **Two categorical**
  Heatmap or mosaic before grouped bars
- **Time series**
  Prefer line, area, slope; avoid bars by default
- **Map data**
  Normalize before choropleths
- **Network**
  Avoid force-directed hairballs with large node counts

::: {.highlight-box}
If the audience's main task is comparison, choose a chart that encodes values by position.
:::

:::

::::

::: {.fragment .highlight-box}
Bookmark these: [data-to-viz.com](https://www.data-to-viz.com) | [r-graph-gallery.com](https://r-graph-gallery.com)
:::


## The Anti-Pie-Chart Argument

```{r}
#| label: fig-pie-vs-bar
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

dat <- tibble(group = LETTERS[1:6],
              value = c(22, 18, 17, 16, 14, 13))

p1 <- ggplot(dat, aes(x = "", y = value, fill = group)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pie chart") +
  theme_void(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p2 <- ggplot(dat, aes(x = value, y = fct_reorder(group, value))) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(aes(label = value), hjust = -0.3, fontface = "bold") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Sorted bar chart", x = NULL, y = NULL) +
  theme_epi553()

p1 + p2 +
  plot_annotation(title = "Pie charts encode angles. Bar charts encode position.",
                  subtitle = "Which one lets you instantly rank the groups?",
                  theme = theme(plot.title = element_text(face = "bold", size = 16),
                                plot.subtitle = element_text(color = "grey40")))
```


# Part 5 {background-color="#46166B" .section-slide}

## The Editorial Style

Cedric Scherer's approach: from default to publication


## The Scherer Approach {.smaller}

:::: {.columns}

::: {.column width="50%"}

::: {.incremental}
1. **Start with the data**, not the chart type
2. **Strip ruthlessly**: remove gridlines, grey backgrounds, redundant titles
3. **Direct label**: put labels next to the lines, not in a legend
4. **Use type as design**: bold title, italic subtitle, grey caption
5. **Annotate**: callouts, arrows, shaded regions
6. **Iterate**: 20+ versions is normal
:::

:::

::: {.column width="50%"}

### The Tools

| Package | Purpose |
|---------|---------|
| `ggtext` | Markdown in titles |
| `patchwork` | Multi-panel figures |
| `ggrepel` | Non-overlapping labels |
| `showtext` | Custom Google Fonts |
| `MetBrewer` | Art-inspired palettes |
| `ggdist` | Uncertainty + distributions |

:::

::::


## Editorial Example: Direct Labels Replace Legends

```{r}
#| label: fig-scherer-style
#| code-fold: true
#| code-summary: "Show the ggplot2 code"
#| fig-height: 5.5

library(ggrepel)

penguin_means <- penguins |>
  drop_na() |>
  summarise(mass = mean(body_mass_g), .by = c(species, year))

ggplot(penguin_means, aes(x = year, y = mass, color = species)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.5) +
  geom_text_repel(
    data = filter(penguin_means, year == max(year)),
    aes(label = species),
    hjust = 0, nudge_x = 0.15, direction = "y",
    segment.color = NA, fontface = "bold", size = 5
  ) +
  scale_color_manual(values = epi_colors) +
  scale_x_continuous(breaks = 2007:2009,
                     expand = expansion(mult = c(0.05, 0.25))) +
  labs(title = "Average Body Mass of Palmer Penguins, 2007-2009",
       subtitle = "Direct labels replace the legend; gridlines are softened",
       x = NULL, y = "Body mass (g)",
       caption = "Source: palmerpenguins R package") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank())
```


# Part 6 {background-color="#46166B" .section-slide}

## A Worked Transformation

Same data. Four versions. Watch the evolution.


## Iteration 1: The Default

```{r}
#| label: fig-iter1
#| fig-height: 5

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  theme_grey(base_size = 14)
```

::: {.muted}
Honest. Also forgettable.
:::


## Iteration 2: Show the Data

```{r}
#| label: fig-iter2
#| fig-height: 5

ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = epi_colors) +
  theme_epi553() +
  theme(legend.position = "none")
```

::: {.fragment .highlight-box}
Now we see how many penguins are in each species and where the outliers actually live.
:::


## Iteration 3: Labels and Theme

```{r}
#| label: fig-iter3
#| fig-height: 5

ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, color = "grey20") +
  geom_jitter(width = 0.2, alpha = 0.4, color = "grey30", size = 1.5) +
  scale_fill_manual(values = epi_colors) +
  labs(title = "Body Mass of Palmer Penguins by Species",
       subtitle = "Boxplots with individual observations overlaid",
       x = NULL, y = "Body mass (g)") +
  theme(legend.position = "none")
```


## Iteration 4: Editorial Polish {.smaller}

```{r}
#| label: fig-iter4
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5.5

library(ggdist)

ggplot(drop_na(penguins, body_mass_g),
       aes(x = species, y = body_mass_g, fill = species)) +
  stat_halfeye(adjust = 0.5, width = 0.6, .width = 0,
               justification = -0.3, point_color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.05, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = epi_colors) +
  coord_cartesian(xlim = c(1.2, NA), clip = "off") +
  labs(title = "Gentoo Penguins Are Substantially Heavier",
       subtitle = "Distribution, median, and individual observations of body mass by species",
       x = NULL, y = "Body mass (g)",
       caption = "Source: palmerpenguins R package") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank())
```

::: {.fragment .finding-box}
The final chart shows the **distribution** (raincloud), the **summary** (boxplot), the **raw data** (jitter), and a **conclusion** baked into the title. That is the difference between a chart and a *finding*.
:::


## Iterations 1 and 2: From Default to Honest

```{r}
#| label: fig-four-iterations-a
#| echo: false
#| fig-height: 5
#| fig-width: 12

p1 <- ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(title = "1. Default", x = NULL, y = NULL) +
  theme_grey(base_size = 13)

p2 <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.4) +
  scale_fill_manual(values = epi_colors) +
  labs(title = "2. Show the data", x = NULL, y = NULL) +
  theme_epi553(base_size = 13) +
  theme(legend.position = "none")

p1 + p2 +
  plot_annotation(
    title = "Step 1: basic summary; Step 2: show the raw data",
    subtitle = "The second version is immediately more informative because students can see spread, outliers, and sample size",
    theme = theme(plot.title = element_text(face = "bold", size = 16),
                  plot.subtitle = element_text(color = "grey40", size = 12))
  )
```

## Iterations 3 and 4: From Clean to Editorial

```{r}
#| label: fig-four-iterations-b
#| echo: false
#| fig-height: 5
#| fig-width: 12

p3 <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, color = "grey20") +
  geom_jitter(width = 0.2, alpha = 0.4, color = "grey30", size = 1.4) +
  scale_fill_manual(values = epi_colors) +
  labs(title = "3. Labels + theme", x = NULL, y = NULL) +
  theme_epi553(base_size = 13) +
  theme(legend.position = "none")

p4 <- ggplot(drop_na(penguins, body_mass_g),
      aes(x = species, y = body_mass_g, fill = species)) +
  stat_halfeye(adjust = 0.5, width = 0.6, .width = 0,
               justification = -0.3, point_color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.05, alpha = 0.3, size = 1) +
  scale_fill_manual(values = epi_colors) +
  coord_cartesian(xlim = c(1.2, NA), clip = "off") +
  labs(title = "4. Editorial polish", x = NULL, y = NULL) +
  theme_epi553(base_size = 13) +
  theme(legend.position = "none", panel.grid.major.x = element_blank())

p3 + p4 +
  plot_annotation(
    title = "Step 3: sharpen the labels; Step 4: add narrative polish",
    subtitle = "Now the audience can compare whether the added design choices actually improve the message",
    theme = theme(plot.title = element_text(face = "bold", size = 16),
                  plot.subtitle = element_text(color = "grey40", size = 12))
  )
```


## Explore These Online {.smaller}

Keep this list short during lecture; share the longer reading list on Brightspace or in your notes.

:::: {.columns}

::: {.column width="50%"}

### Start here

- [From Data to Viz](https://www.data-to-viz.com)
- [R Graph Gallery](https://r-graph-gallery.com)
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)

:::

::: {.column width="50%"}

### For inspiration

- [Cedric Scherer's portfolio](https://www.cedricscherer.com/top/dataviz/)
- [The Pudding](https://pudding.cool)
- [Our World in Data](https://ourworldindata.org)

:::

::::


## Foundations Summary {.smaller}

:::: {.columns}

::: {.column width="50%"}

- **Build in layers**
  Start with data, mapping, and geometry
- **Show the data**
  Prefer raw observations plus a summary
- **Use strong encodings**
  Position usually beats area and color
- **Quantify uncertainty**
  CIs and distribution displays matter

:::

::: {.column width="50%"}

- **Label directly**
  Reduce legend lookups when you can
- **Strip clutter**
  Remove anything that does not aid comparison
- **Write takeaway titles**
  Titles should state the finding
- **Iterate**
  Your fourth version is often the best one

:::

::::

# Part 7 {background-color="#46166B" .section-slide}

## Color, Type, and Accessibility

Choosing color is a design *and* an ethics decision.


## Color Principles {.smaller}

:::: {.columns}

::: {.column width="55%"}

### The Four Types of Color Scales

1. [**Categorical / Qualitative**]{.emphasis}: distinct hues, no ranking
   - `Set2`, `Dark2`, Okabe-Ito
2. [**Sequential**]{.emphasis}: light to dark, one or two hues
   - `viridis`, `mako`, `Blues`
3. [**Diverging**]{.emphasis}: two hues meeting at a neutral midpoint
   - `RdBu`, `BrBG` (only when there is a meaningful zero)
4. [**Highlight**]{.emphasis}: one color for the story, grey for context

:::

::: {.column width="45%"}

```{r}
#| label: fig-okabe-ito
#| echo: false
#| fig-height: 6

okabe_ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
               "#0072B2", "#D55E00", "#CC79A7", "#000000")

tibble(
  color = okabe_ito,
  name = c("Orange", "Sky Blue", "Green", "Yellow",
           "Blue", "Vermillion", "Pink", "Black"),
  y = rev(seq_along(okabe_ito))
) |>
  ggplot(aes(x = 1, y = y, fill = color)) +
  geom_tile(width = 0.6, height = 0.85) +
  geom_text(aes(label = paste(name, " ", color)),
            hjust = -0.1, x = 1.35, size = 5, fontface = "bold") +
  scale_fill_identity() +
  scale_x_continuous(limits = c(0.5, 3.5)) +
  labs(title = "Okabe-Ito Palette",
       subtitle = "Designed for colorblind accessibility") +
  theme_void(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))
```

:::

::::


## Why Colorblind Safety Matters

About **8% of men** and **0.5% of women** have some form of color vision deficiency.

```{r}
#| label: fig-colorblind-demo
#| echo: false
#| fig-height: 4.5

p_bad <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("red", "green", "blue")) +
  labs(title = "Red-Green-Blue: invisible to ~8% of men",
       x = "Bill length (mm)", y = "Bill depth (mm)") +
  theme_epi553(base_size = 12)

p_good <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
  labs(title = "Okabe-Ito: safe for all viewers",
       x = "Bill length (mm)", y = "Bill depth (mm)") +
  theme_epi553(base_size = 12)

p_bad + p_good +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

::: {.highlight-box}
**Rule of thumb**: use `viridis`-family palettes by default. Check with `colorBlindness::cvdPlot()` or [coblis.com](https://www.color-blindness.com/coblis-color-blindness-simulator/).
:::


## Typography and Direct Labeling

:::: {.columns}

::: {.column width="50%"}

### Typography Rules

- **One font family** consistently
- Title: **bold, slightly larger**
- Subtitle: regular, grey
- Caption: small, grey, data source
- Avoid serifs in dense charts
- Good sans-serifs: Atkinson Hyperlegible, Inter, Source Sans Pro

:::

::: {.column width="50%"}

### Direct Labeling

Put the label **next to the thing it labels**.

Legends force the eye to bounce between chart and key.

Direct labels eliminate that bounce.

::: {.fragment}
```r
# Instead of a legend:
geom_text_repel(
  data = filter(df, year == max(year)),
  aes(label = group)
)
```
:::

:::

::::


## The Visualization Checklist {.smaller}

::: {.checklist}

Before you submit a chart, ask:

- Does the title state the **finding**, not the topic?
- Is the y-axis at zero (for bar charts) or appropriately scaled?
- Are **units** labeled?
- Is there a **source caption**?
- Is the color palette **colorblind-safe**?
- Are uncertainty intervals shown?
- Could a reader explain the chart in **one sentence**?

:::


# Part 8 {background-color="#46166B" .section-slide}

## Modern Techniques and Advanced Aesthetics

The "show me what is possible" tour.


## 8.1 Highlighting with gghighlight {.smaller}

Many groups, but only a few matter to your story. **Highlighting** fades out the context.

:::: {.columns}

::: {.column width="45%"}

```{r}
#| label: fig-highlight-code
#| eval: false

library(gghighlight)

gapminder_like <- penguins |>
  drop_na() |>
  summarise(
    mean_mass = mean(body_mass_g),
    .by = c(species, year)
  )

ggplot(gapminder_like,
       aes(x = year, y = mean_mass,
           color = species)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.5) +
  gghighlight(
    species == "Gentoo",
    unhighlighted_params = list(
      color = "grey80",
      linewidth = 0.8)) +
  scale_color_manual(
    values = c("Gentoo" = "#2E86AB"))
```

:::

::: {.column width="55%"}

```{r}
#| label: fig-highlight
#| echo: false
#| fig-height: 6

library(gghighlight)

gapminder_like <- penguins |>
  drop_na() |>
  summarise(mean_mass = mean(body_mass_g), .by = c(species, year))

ggplot(gapminder_like, aes(x = year, y = mean_mass, color = species)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.5) +
  gghighlight(species == "Gentoo",
              unhighlighted_params = list(color = "grey80", linewidth = 0.8)) +
  scale_color_manual(values = c("Gentoo" = "#2E86AB")) +
  labs(title = "Gentoo Penguins Stand Out",
       subtitle = "Other species recede into the background",
       x = NULL, y = "Mean body mass (g)") +
  theme(legend.position = "none")
```

:::

::::

::: {.highlight-box}
The eye is drawn instantly to the highlighted line, but the reader still has full context.
:::


## 8.2 Animation with gganimate {.smaller}

Time-series and longitudinal data come alive when animated. Build a standard ggplot, then add a `transition_*()` layer.

```{r}
#| eval: false

library(gganimate)
library(gapminder)

p <- ggplot(gapminder,
            aes(x = gdpPercap, y = lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_size(range = c(2, 12), guide = "none") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Year: {frame_time}",
       x = "GDP per capita (log scale)", y = "Life expectancy (years)") +
  theme_epi553() +
  transition_time(year) +        # <-- this is the only new line
  ease_aes("cubic-in-out")

animate(p, nframes = 100, fps = 10, width = 800, height = 500)
```

### Key Transition Functions

| Function | What it animates | Example |
|----------|------------------|---------|
| `transition_time()` | Continuous variable (year, day) | Gapminder bubble chart |
| `transition_states()` | Discrete variable, pauses on each | Before/after comparison |
| `transition_reveal()` | Progressive reveal along an axis | Drawing a time series |

::: {.highlight-box}
**Tip from Cedric Scherer:** animations should serve a narrative purpose. If the same insight is clearer in a static small-multiple, prefer the static version.
:::


## 8.3 Interactive Charts {.smaller}

### plotly: instant interactivity from any ggplot

```{r}
#| label: fig-plotly
#| fig-height: 4.5

library(plotly)

p <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g,
                          color = species,
                          text = paste0("Island: ", island,
                                        "<br>Sex: ", sex,
                                        "<br>Year: ", year))) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = epi_colors) +
  labs(x = "Bill length (mm)", y = "Body mass (g)") +
  theme_epi553(base_size = 11)

ggplotly(p, tooltip = "text", height = 400)
```


## 8.3 Interactive Charts: ggiraph {.smaller}

ggiraph: finer control, hover effects, click events. **Hover over points below!**

```{r}
#| label: fig-ggiraph
#| code-fold: true
#| code-summary: "Show the ggiraph code"

library(ggiraph)

p <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g, color = species)) +
  geom_point_interactive(
    aes(tooltip = paste("Species:", species, "<br>Island:", island,
                        "<br>Mass:", body_mass_g, "g"),
        data_id = species),
    size = 2.5, alpha = 0.7
  ) +
  scale_color_manual(values = epi_colors) +
  labs(x = "Bill length (mm)", y = "Body mass (g)") +
  theme_epi553(base_size = 11)

girafe(ggobj = p, height_svg = 3.5, width_svg = 8,
       options = list(
         opts_hover(css = "stroke:black;stroke-width:2px;fill-opacity:1;"),
         opts_hover_inv(css = "opacity:0.2;")
       ))
```


## 8.4 Distributions: Raincloud Plots {.smaller}

Boxplots hide bimodality. Histograms hide group differences. **Raincloud plots show everything.**

```{r}
#| label: fig-raincloud
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

library(ggdist)

ggplot(drop_na(penguins, body_mass_g),
       aes(x = body_mass_g, y = species, fill = species)) +
  stat_halfeye(adjust = 0.6, .width = 0, justification = -0.2,
               point_color = NA, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(height = 0.07, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = epi_colors) +
  labs(title = "Raincloud Plot of Penguin Body Mass",
       subtitle = "Distribution + boxplot + raw observations in one chart",
       x = "Body mass (g)", y = NULL) +
  theme(legend.position = "none")
```


## 8.4 Distributions: Ridge Plots {.smaller}

When you want to compare distributions across **many groups** in a small space.

```{r}
#| label: fig-ridges
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

library(ggridges)

ggplot(diamonds, aes(x = price, y = cut, fill = cut)) +
  geom_density_ridges(alpha = 0.8, scale = 1.1) +
  scale_x_log10(labels = label_dollar()) +
  scale_fill_viridis_d(option = "mako") +
  labs(title = "Distribution of Diamond Prices by Cut",
       subtitle = "Ridge plots make many distributions easy to compare",
       x = "Price (log scale)", y = NULL) +
  theme(legend.position = "none")
```


## 8.5 Uncertainty You Can See {.smaller}

Confidence intervals as skinny error bars often get missed. Show the **whole distribution**.

```{r}
#| label: fig-uncertainty
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

library(ggdist)
library(distributional)

estimates <- tibble(
  predictor = c("Smoking", "Exercise", "Income", "Sleep", "Age"),
  estimate  = c(0.65, -0.45, -0.20, -0.30, 0.05),
  se        = c(0.10, 0.08, 0.07, 0.09, 0.04)
)

ggplot(estimates, aes(y = fct_reorder(predictor, estimate),
                      xdist = dist_normal(estimate, se))) +
  stat_halfeye(.width = c(0.5, 0.95), fill = "#2E86AB",
               slab_alpha = 0.6, point_size = 3.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(title = "Coefficient Plot with Full Sampling Distributions",
       subtitle = "Thick band = 50% interval, thin line = 95% interval",
       x = "Log odds ratio", y = NULL)
```

::: {.fragment .finding-box}
This is far more honest than a forest plot of point estimates with whiskers, because the reader can see the **shape** of the uncertainty.
:::


## 8.6 Patchwork: Composing Multi-Panel Figures {.smaller}

```{r}
#| label: fig-patchwork
#| code-fold: true
#| code-summary: "Show the patchwork code"
#| fig-height: 5.5

p_scatter <- ggplot(penguins, aes(bill_length_mm, body_mass_g, color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = epi_colors) +
  theme(legend.position = "none") +
  labs(x = "Bill length (mm)", y = "Body mass (g)")

p_top <- ggplot(penguins, aes(bill_length_mm, fill = species)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = epi_colors) +
  theme_void() + theme(legend.position = "none")

p_right <- ggplot(penguins, aes(body_mass_g, fill = species)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = epi_colors) +
  coord_flip() +
  theme_void() + theme(legend.position = "none")

(p_top + plot_spacer() + p_scatter + p_right +
    plot_layout(ncol = 2, widths = c(4, 1), heights = c(1, 4))) +
  plot_annotation(
    title = "Marginal Density + Scatterplot",
    subtitle = "Composed with patchwork: p_top + plot_spacer() + p_scatter + p_right",
    theme = theme(plot.title = element_text(face = "bold", size = 16),
                  plot.subtitle = element_text(color = "grey40", family = "Fira Code",
                                               size = 11))
  )
```


## Patchwork Syntax Cheat Sheet

The arithmetic is intuitive:

```{r}
#| eval: false
#| code-line-numbers: "1|2|3|4|5"

p1 + p2              # side by side
p1 / p2              # stacked vertically
(p1 | p2) / p3       # two on top, one below
p1 + p2 + p3 + plot_layout(ncol = 2)  # grid layout
p1 + plot_annotation(title = "Combined Figure")  # annotation
```


## 8.7 Annotation as a First-Class Citizen {.smaller}

The fastest way to make a chart "editorial": **annotate the finding directly on the chart**.

```{r}
#| label: fig-annotation
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

library(ggtext)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7, size = 2.2) +
  annotate("curve", x = 175, y = 5500, xend = 210, yend = 5100,
           arrow = arrow(length = unit(0.25, "cm")),
           curvature = -0.3, color = "grey30", linewidth = 0.6) +
  annotate("label", x = 173, y = 5650,
           label = "Gentoos cluster in the\nheavy / long-flipper corner",
           hjust = 0, size = 4.2, color = "grey20", lineheight = 0.95,
           fill = "#fff8e1", label.size = 0, fontface = "italic") +
  scale_color_manual(values = epi_colors) +
  labs(title = "Body mass scales with flipper length, but <span style='color:#46166B;'>**species matter**</span>",
       subtitle = "Annotated with a callout arrow and label box",
       x = "Flipper length (mm)", y = "Body mass (g)",
       color = NULL) +
  theme(plot.title = element_markdown(face = "bold", size = 16),
        legend.position = "top")
```


## 8.8 Spatial Visualization with sf {.smaller}

Public health data are almost always spatial. The `sf` package gives ggplot2 first-class support.

```{r}
#| label: fig-spatial
#| code-fold: true
#| code-summary: "Show the spatial code"
#| fig-height: 5.5

library(sf)
library(tigris)
options(tigris_use_cache = TRUE)

ny_counties <- counties(state = "NY", cb = TRUE, class = "sf")
ny_counties$fmd_prev <- runif(nrow(ny_counties), 8, 18)

ggplot(ny_counties) +
  geom_sf(aes(fill = fmd_prev), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "rocket", direction = -1,
                       name = "FMD %") +
  labs(title = "Frequent Mental Distress Prevalence by NY County",
       subtitle = "Hypothetical data for demonstration",
       caption = "Source: U.S. Census Bureau (TIGER/Line)") +
  theme_void(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))
```


## 8.9 Tables as Visualizations: gt + gtExtras {.smaller}

Sometimes the right visualization is a **table**. `gt` makes editorial-quality tables with inline plots.

```{r}
#| label: tbl-gt-demo
#| code-fold: true
#| code-summary: "Show the gt code"

library(gt)
library(gtExtras)

penguins |>
  drop_na() |>
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g),
    sd_mass = sd(body_mass_g),
    masses = list(body_mass_g),
    .by = species
  ) |>
  gt() |>
  gt_plt_dist(masses, type = "density", fill_color = "#2E86AB") |>
  fmt_number(mean_mass, decimals = 0) |>
  fmt_number(sd_mass, decimals = 0) |>
  cols_label(species = "Species", n = "N",
             mean_mass = "Mean (g)", sd_mass = "SD (g)",
             masses = "Distribution") |>
  tab_header(title = md("**Penguin Body Mass by Species**"),
             subtitle = "Summary statistics with inline density plots") |>
  gt_theme_538() |>
  tab_options(table.width = pct(90))
```


## 8.10 Heatmaps and Correlation Matrices {.smaller}

Heatmaps turn a matrix of numbers into a pattern you can *see*. Essential for correlation tables, confusion matrices, and time-by-group summaries.

```{r}
#| label: fig-heatmap
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5.5

# Correlation matrix of numeric penguin variables
cor_data <- penguins |>
  drop_na() |>
  select(where(is.numeric)) |>
  cor() |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "cor")

ggplot(cor_data, aes(x = var1, y = var2, fill = cor)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(cor, 2),
                color = abs(cor) > 0.6),
            size = 4.5, fontface = "bold") +
  scale_fill_gradient2(low = "#D55E00", mid = "white", high = "#0072B2",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Correlation") +
  scale_color_manual(values = c("TRUE" = "white", "FALSE" = "grey20"),
                     guide = "none") +
  scale_x_discrete(labels = \(x) str_replace_all(x, "_", "\n")) +
  scale_y_discrete(labels = \(x) str_replace_all(x, "_", "\n")) +
  labs(title = "Correlation Heatmap of Penguin Measurements",
       subtitle = "Color intensity encodes strength; text encodes exact value",
       x = NULL, y = NULL) +
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        panel.grid = element_blank())
```


## 8.11 Dumbbell Charts: Showing Change {.smaller}

When you need to show the difference between two time points or conditions, **dumbbell charts** are more effective than grouped bars.

```{r}
#| label: fig-dumbbell
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

# Simulated epi example: disease rates before/after intervention
intervention_data <- tibble(
  county = c("Albany", "Saratoga", "Rensselaer", "Schenectady",
             "Columbia", "Greene", "Warren", "Washington"),
  before = c(15.2, 12.8, 18.1, 16.5, 11.3, 14.7, 9.8, 13.2),
  after  = c(11.1, 10.2, 12.4, 13.8, 9.5, 11.0, 8.1, 10.9)
) |>
  mutate(change = after - before,
         county = fct_reorder(county, change))

ggplot(intervention_data) +
  geom_segment(aes(x = before, xend = after,
                   y = county, yend = county),
               color = "grey60", linewidth = 1.2) +
  geom_point(aes(x = before, y = county), color = "#D55E00",
             size = 4) +
  geom_point(aes(x = after, y = county), color = "#0072B2",
             size = 4) +
  annotate("text", x = 19, y = 8.3, label = "Before",
           color = "#D55E00", fontface = "bold", size = 4.5) +
  annotate("text", x = 19, y = 7.7, label = "After",
           color = "#0072B2", fontface = "bold", size = 4.5) +
  labs(title = "Every County Improved After the Intervention",
       subtitle = "Rate per 1,000 population, before vs. after community health program",
       x = "Rate per 1,000", y = NULL) +
  scale_x_continuous(limits = c(7, 20))
```

::: {.highlight-box}
Dumbbell charts encode **direction**, **magnitude**, and **rank** simultaneously. Far more effective than grouped bar charts for before/after comparisons.
:::


## 8.12 Build Your Own Reusable Theme {.smaller}

Package your style as a function. Reuse it across every chart.

```{r}
#| code-line-numbers: "1-2|3-5|6-8|9-12|13-15"

theme_epi553 <- function(base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 4,
                                color = "#1a1a2e"),
      plot.subtitle = element_text(color = "grey40"),
      plot.caption = element_text(color = "grey60", size = base_size - 3,
                                  hjust = 0),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      axis.title = element_text(color = "grey30"),
      strip.text = element_text(face = "bold"),
      legend.position = "top",
      plot.title.position = "plot",
      plot.caption.position = "plot"
    )
}
```

::: {.highlight-box}
**Practical tip**: drop `theme_epi553()` into a `themes.R` file. Source it from every analysis. Consistent figures without effort.
:::


## 8.13 The Modern Workflow

::: {.incremental}

1. **Sketch first.** On paper. What is the story?
2. **Prototype in ggplot2** with default themes. Get the geometry right.
3. **Iterate the encoding.** Try three chart types. Pick one.
4. **Layer in annotations.** Title is the finding. Direct labels. Callouts.
5. **Polish the theme.** Fonts, colors, spacing.
6. **Export at the right resolution.**

:::

::: {.fragment}
```{r}
#| eval: false

ggsave("figure_1.png", width = 8, height = 5, dpi = 300, bg = "white")
ggsave("figure_1.svg", width = 8, height = 5)        # vector for editorial
ggsave("figure_1.pdf", width = 8, height = 5,
       device = cairo_pdf)                             # high-quality PDF
```
:::

::: {.fragment .finding-box}
**Step 7:** Show it to a colleague. If they cannot explain the chart in 10 seconds, iterate.
:::


# Part 9 {background-color="#46166B" .section-slide}

## Visualizing Regression Models

You have spent the semester building models. Now make them *visible*.


## The Problem with Regression Tables {.smaller}

:::: {.columns}

::: {.column width="45%"}

### What reviewers see

A wall of numbers:

| Term | Estimate | SE | p |
|------|----------|-----|------|
| (Intercept) | -1.23 | 0.41 | 0.003 |
| Smoking | 0.65 | 0.10 | <0.001 |
| Exercise | -0.45 | 0.08 | <0.001 |
| Income | -0.20 | 0.07 | 0.004 |
| Sleep | -0.30 | 0.09 | 0.001 |
| Age | 0.05 | 0.04 | 0.211 |

:::

::: {.column width="55%"}

### What readers *understand*

```{r}
#| label: fig-coef-plot-intro
#| echo: false
#| fig-height: 5.5

library(ggdist)
library(distributional)

estimates <- tibble(
  predictor = c("Smoking", "Exercise", "Income", "Sleep", "Age"),
  estimate  = c(0.65, -0.45, -0.20, -0.30, 0.05),
  se        = c(0.10, 0.08, 0.07, 0.09, 0.04)
) |>
  mutate(
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se,
    significant = !(lower <= 0 & upper >= 0)
  )

ggplot(estimates, aes(x = estimate,
                      y = fct_reorder(predictor, estimate),
                      color = significant)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  size = 0.8, linewidth = 1.1) +
  scale_color_manual(values = c("TRUE" = "#2E86AB", "FALSE" = "grey60"),
                     guide = "none") +
  labs(title = "Same Model, Instantly Readable",
       subtitle = "Blue = significant at p < 0.05; grey = non-significant",
       x = "Log odds ratio (95% CI)", y = NULL)
```

:::

::::

::: {.finding-box}
A coefficient plot communicates **direction**, **magnitude**, **uncertainty**, and **significance** in one glance. A table requires row-by-row mental math.
:::


## Forest Plots from Real Models {.smaller}

Fit a logistic regression on the penguins data and plot the odds ratios directly.

```{r}
#| label: fig-forest-real
#| code-fold: true
#| code-summary: "Show the full code"
#| fig-height: 5.5

library(broom)

# Fit a logistic regression: predict heavy penguin (above median mass)
model_data <- penguins |>
  drop_na() |>
  mutate(heavy = as.integer(body_mass_g > median(body_mass_g)))

fit <- glm(heavy ~ bill_length_mm + bill_depth_mm + flipper_length_mm +
             species + sex,
           data = model_data, family = binomial)

# Tidy the model output (use Wald CIs for stability)
model_tidy <- tidy(fit, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    conf.low = exp(log(estimate) - 1.96 * std.error),
    conf.high = exp(log(estimate) + 1.96 * std.error),
    term = case_match(term,
      "bill_length_mm" ~ "Bill length (mm)",
      "bill_depth_mm" ~ "Bill depth (mm)",
      "flipper_length_mm" ~ "Flipper length (mm)",
      "speciesChinstrap" ~ "Chinstrap vs. Adelie",
      "speciesGentoo" ~ "Gentoo vs. Adelie",
      "sexmale" ~ "Sex (male vs. female)"
    ),
    significant = p.value < 0.05
  )

ggplot(model_tidy, aes(x = estimate, y = fct_reorder(term, estimate),
                       color = significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  size = 0.8, linewidth = 1.1) +
  geom_text(aes(label = sprintf("OR = %.2f", estimate)),
            vjust = -1, size = 3.8, show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "#e64173", "FALSE" = "grey60"),
                     guide = "none") +
  scale_x_log10() +
  labs(title = "Predictors of Above-Median Body Mass (Logistic Regression)",
       subtitle = "Odds ratios with 95% CI on log scale; red = p < 0.05",
       x = "Odds Ratio (log scale)", y = NULL,
       caption = "Model: glm(heavy ~ bill + flipper + species + sex, family = binomial)")
```


## Predicted Probability Curves {.smaller}

Show what your model *predicts* across the range of a key variable, holding others at their means.

```{r}
#| label: fig-predicted-probs
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

library(broom)

# Generate predictions across flipper length range
pred_grid <- tibble(
  flipper_length_mm = seq(170, 235, length.out = 200),
  bill_length_mm = mean(model_data$bill_length_mm),
  bill_depth_mm = mean(model_data$bill_depth_mm),
  species = "Adelie",
  sex = "female"
)

preds <- augment(fit, newdata = pred_grid, type.predict = "response",
                 se_fit = TRUE) |>
  mutate(lower = pmax(.fitted - 1.96 * .se.fit, 0),
         upper = pmin(.fitted + 1.96 * .se.fit, 1))

ggplot(preds, aes(x = flipper_length_mm, y = .fitted)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "#2E86AB", alpha = 0.2) +
  geom_line(color = "#2E86AB", linewidth = 1.3) +
  geom_rug(data = model_data,
           aes(x = flipper_length_mm, y = heavy),
           sides = "tb", alpha = 0.15, color = "grey40") +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Predicted Probability of Above-Median Mass by Flipper Length",
       subtitle = "Logistic regression (Adelie, female); rug marks show observed data",
       x = "Flipper length (mm)", y = "Predicted probability",
       caption = "Shaded band = approximate 95% confidence interval")
```

::: {.highlight-box}
Predicted probability curves are one of the most effective ways to communicate logistic regression results to non-statisticians.
:::


## Marginal Effects with ggeffects {.smaller}

The `ggeffects` package automates predicted value plots for any model class.

:::: {.columns}

::: {.column width="45%"}

```{r}
#| label: fig-ggeffects-code
#| eval: false

library(ggeffects)

# One line to get predicted values
preds <- ggpredict(fit,
  terms = c("flipper_length_mm",
            "sex"))

# Built-in plot method
plot(preds) +
  labs(
    title = "Marginal Effect of Flipper Length by Sex"
  )

# Or extract data for custom ggplot
as.data.frame(preds) |>
  ggplot(aes(x, predicted,
             color = group)) +
  geom_ribbon(...) +
  geom_line(...)
```

:::

::: {.column width="55%"}

```{r}
#| label: fig-ggeffects
#| echo: false
#| fig-height: 6

library(ggeffects)

preds <- ggpredict(fit, terms = c("flipper_length_mm [170:235 by=1]", "sex"),
                   condition = c(species = "Adelie"))

ggplot(as.data.frame(preds),
       aes(x = x, y = predicted, color = group, fill = group)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15,
              color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("female" = "#D55E00", "male" = "#0072B2")) +
  scale_fill_manual(values = c("female" = "#D55E00", "male" = "#0072B2")) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Marginal Effect of Flipper Length by Sex (Adelie)",
       subtitle = "ggeffects automates predicted values for any model",
       x = "Flipper length (mm)", y = "P(above-median mass)",
       color = "Sex", fill = "Sex")
```

:::

::::


## Diagnostic Plots: the `performance` + `see` Packages {.smaller}

The `performance` package (from easystats) provides model diagnostics; `see` visualizes them with ggplot2.

```{r}
#| label: fig-diagnostics
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

# Fit a linear model for diagnostics demo
lm_fit <- lm(body_mass_g ~ bill_length_mm + bill_depth_mm + flipper_length_mm + species,
             data = drop_na(penguins))

library(performance)
library(see)

check_model(lm_fit, check = c("linearity", "normality", "qq", "homogeneity"))
```

::: {.highlight-box}
`check_model()` replaces the base R `plot(model)` with a modern, multi-panel diagnostic dashboard. One function call, publication-ready output.
:::


## Model Comparison Visualization {.smaller}

When comparing nested or competing models, visualize the fit statistics side by side.

```{r}
#| label: fig-model-comparison
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

# Fit competing models
m1 <- glm(heavy ~ flipper_length_mm, data = model_data, family = binomial)
m2 <- glm(heavy ~ flipper_length_mm + species, data = model_data, family = binomial)
m3 <- glm(heavy ~ flipper_length_mm + species + sex, data = model_data, family = binomial)
m4 <- fit  # full model from earlier

# Helper to get Wald CIs
tidy_wald <- function(mod, label) {
  tidy(mod, exponentiate = TRUE) |>
    mutate(conf.low = exp(log(estimate) - 1.96 * std.error),
           conf.high = exp(log(estimate) + 1.96 * std.error),
           model = label)
}

# Compare coefficients across models
models_tidy <- bind_rows(
  tidy_wald(m1, "Model 1:\nFlipper only"),
  tidy_wald(m2, "Model 2:\n+ Species"),
  tidy_wald(m3, "Model 3:\n+ Sex"),
  tidy_wald(m4, "Model 4:\n+ Bill measures")
) |>
  filter(term == "flipper_length_mm")

ggplot(models_tidy, aes(x = estimate, y = model)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  color = "#2E86AB", size = 1, linewidth = 1.1) +
  geom_text(aes(label = sprintf("OR = %.2f (%.2f, %.2f)",
                                estimate, conf.low, conf.high)),
            vjust = -1.2, size = 3.8, color = "grey30") +
  scale_x_log10() +
  labs(title = "How Stable Is the Flipper Length Effect Across Models?",
       subtitle = "Odds ratio for flipper_length_mm as covariates are added",
       x = "Odds Ratio (log scale)", y = NULL,
       caption = "Stable estimates across nested models suggest robust association")
```


## The #30DayChartChallenge {.smaller}

Every **April**, the data visualization community participates in the [#30DayChartChallenge](https://github.com/30DayChartChallenge/): one prompt per day, one chart per day, shared on social media. Created in 2021 by **Cedric Scherer** and **Dominic Roye**, inspired by the #30DayMapChallenge.

:::: {.columns}

::: {.column width="50%"}

### Why It Matters for You

::: {.incremental}
- Forces you to **try chart types you would never pick** (waffle charts, bump charts, slope graphs, treemaps)
- Builds a **public portfolio** of your work
- Exposes you to how the **global community** solves the same prompt differently
- Many prompts are epi-relevant: *uncertainty*, *distributions*, *time series*, *part-to-whole*, *relationships*
:::

### Five Categories (based on "The Graphic Continuum")

1. **Comparisons** (days 1-6)
2. **Distributions** (days 7-12)
3. **Relationships** (days 13-18)
4. **Time series** (days 19-24)
5. **Uncertainties** (days 25-30)

::: {.muted}
Participants choose their own data and tools freely. The 2026 edition is at [github.com/30DayChartChallenge/Edition2026](https://github.com/30DayChartChallenge/Edition2026).
:::

:::

::: {.column width="50%"}

### Techniques You Can Steal

| Prompt | Chart Type | R Package |
|--------|-----------|-----------|
| Part-to-whole | Waffle chart | `waffle` |
| Ranking | Bump chart | `ggbump` |
| Slope | Slope chart | `geom_segment` |
| Circular | Polar bar | `coord_polar()` |
| Uncertainty | Gradient intervals | `ggdist` |
| Relationships | Network | `ggraph` + `tidygraph` |
| Neo-geometric | Voronoi | `ggforce` |
| Storytelling | Annotated timeline | `ggtext` + `annotate` |

### Where to Explore

- [github.com/30DayChartChallenge](https://github.com/30DayChartChallenge/) -- all editions since 2021
- Search `#30DayChartChallenge` on Twitter/X or Mastodon
- [Cedric Scherer's contributions](https://github.com/z3tt/30DayChartChallenge) -- R code for every entry
- [R Graph Gallery](https://r-graph-gallery.com) -- reproducible R code for all chart types

:::

::::


## Bonus: Waffle Chart {.smaller}

A waffle chart is a **part-to-whole** alternative to pie charts, popularized by the #30DayChartChallenge. Each square = 1 unit.

```{r}
#| label: fig-waffle
#| code-fold: true
#| code-summary: "Show the waffle code"
#| fig-height: 4.5

library(waffle)

penguin_counts <- penguins |>
  drop_na() |>
  count(species) |>
  mutate(n_scaled = round(n / 5))  # each square = 5 penguins

waffle(
  c("Adelie" = penguin_counts$n_scaled[1],
    "Chinstrap" = penguin_counts$n_scaled[2],
    "Gentoo" = penguin_counts$n_scaled[3]),
  rows = 5,
  size = 1,
  colors = c("#FF6B35", "#A23B72", "#2E86AB"),
  title = "Palmer Penguins by Species",
  xlab = "1 square = 5 penguins"
) +
  theme(plot.title = element_text(face = "bold", size = 16),
        legend.position = "bottom")
```

::: {.highlight-box}
**Challenge for you:** pick a prompt from the [#30DayChartChallenge](https://github.com/30DayChartChallenge/) and create a chart using a dataset from this semester. Share it!
:::


## Lecture Summary {.smaller}

:::: {.columns}

::: {.column width="50%"}

- **Build plots in layers**
- **Show raw data whenever possible**
- **Use uncertainty displays, not just point estimates**
- **Prefer position over area for comparison**
- **Make titles state the conclusion**

:::

::: {.column width="50%"}

- **Use colorblind-safe palettes by default**
- **Direct-label when the audience benefits**
- **Remove nonessential visual clutter**
- **Visualize models, do not just tabulate them**
- **Iterate until the story is obvious**

:::

::::


## References {.smaller}

:::: {.columns}

::: {.column width="50%"}

### Books

- Healy, K. (2018). *Data Visualization: A Practical Introduction*. [socviz.co](https://socviz.co)
- Wilke, C. O. (2019). *Fundamentals of Data Visualization*. [clauswilke.com/dataviz](https://clauswilke.com/dataviz/)
- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. [ggplot2-book.org](https://ggplot2-book.org)

### Foundational

- Tufte, E. (2001). *The Visual Display of Quantitative Information* (2nd ed.)
- Cleveland, W. S., & McGill, R. (1984). Graphical perception. *JASA*, 79(387), 531-554.

:::

::: {.column width="50%"}

### Online Resources

- Holtz, Y. [R Graph Gallery](https://r-graph-gallery.com)
- Holtz, Y. & Healy, C. [From Data to Viz](https://www.data-to-viz.com)
- Scherer, C. [Portfolio](https://www.cedricscherer.com)

### Next Lecture (April 28)

Course Review: putting the entire semester together.

:::

::::


# {background-color="#0e1c36"}

::: {style="text-align: center; margin-top: 20%;"}

[Thank You]{style="font-family: 'Playfair Display'; font-size: 2.5em; color: #EAAA00; font-weight: 800;"}

[EPI 553 | University at Albany]{style="color: rgba(255,255,255,0.7); font-size: 0.8em;"}

:::
